---
title: "A Full Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Full Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## A Full Workflow in Bioconductor

In this document, we run a full analysis on our `RangedSummarizedExperiment` by prerprocessing and running `DESeq()` on the data. We then plot our candidates using the `EnhancedVolcano` package.

```{r}
library(SummarizedExperiment)
library(tidySummarizedExperiment)
library(tidyverse)
library(DESeq2)
library(EnhancedVolcano)

GSE96870 <- readRDS(system.file("extdata/GSE96870_se.rds", package="bioc2026Intro"))
GSE96870
```
 
## Preprocessing and Filtering

We first need to convert our `RangedSummarizedExperiment` object into a `DESeq2` object, which we do by specifying the design formula.

Then we can filter out genes by low total expression counts (5 counts or less).

```{r}
GSE96870_deseq <- 
  DESeqDataSet(GSE96870, design = ~ sex + time)  #make DESeqDataset

GSE96870_filtered <- GSE96870_deseq[rowSums(assay(GSE96870_deseq)) > 5,] #filter out low expressing candidates
```


## Run DESeq

The `DESeq()` function will preprocess our data by estimating the size factors of each sample, allowing us to normalize each sample by library size. The actual modeling of the expression candidates is done by `nbinomialWaldTest`.

```{r}
GSE96870_fit <- DESeq(GSE96870_filtered) #run estimateSizeFactors, estimateDispersions, nbinomialWaldTest
```

## Filter Results by Alpha, Visualize

We use the `results()` function in `DESeq2` to filter our candidates using an alpha threshold of 0.05 and a log-fold change threshold of 1 (which corresponds to a 2 fold expression difference between our two conditions, "Day0", and "Day8").

Finally, we can take those candidates and then plot a volcano plot to visualize -log2 fold-change versus -log10 p-value.

```{r}
GSE_results <- results(GSE96870_fit, 
                       contrast = c("time", "Day8", "Day0"), 
                       lfcThreshold = 1,
                       alpha = 0.05) #calculate results from the contrast

GSE_results

EnhancedVolcano(GSE_results,  
                lab = rownames(GSE_results),
                x = 'log2FoldChange',
                y = 'pvalue')
```



